{\small} CUBEP3M: High Performance P$^{3}$M N-body code

### 2012-05-01 Extendedpp.tex
Upgraded extendedpp.tex, plus minor corrections everywhere.

 To quantify the accuracy improvement versus computing time requirements, we performed the following test.
 We generate a set of initial conditions at a starting redshift of $z = XXX$, with a box size equal to $ XXX h^{-1}\mbox{Mpc}$,
 and with $XXX^{3}$ particles. We evolve the particles to $z=0$ with different ranges for the pp calculation, and compare 
 the resulting power spectra. For the results to be meaningful, we also need to use the same random seed for the random number generator,
 such that the only difference between different runs is the range of the pp force.
{\bf (JD, insert here your P(k) plot and a description of what you observe, in terms of accuracy improvement and real time cpu usage...)}

>>>>

 To quantify the accuracy improvement versus computing time requirements, we performed the following test.
 We generate a set of initial conditions at a starting redshift of $z = 100$, with a box size equal to $ 200 h^{-1}\mbox{Mpc}$,
 and with $128^{3}$ particles. We evolve the particles to $z=0$ with different ranges for the pp calculation, and compare 
 the resulting power spectra. For the results to be meaningful, we also need to use the same random seed for the random number generator,
 such that the only difference between different runs is the range of the pp force.
 Fig. \ref{fig:power} shows the dimensionless power spectrum of the different runs, where we see a significant gains in resolution
 when extending  PM to P$^{3}$M first, and when adding successive layers of fine cell mesh where the pp force is extended.
We have not plotted the results for higher numbers of layers, as the improvement becomes milder there: the fine grid calculations
are more accurate as the distance increases. For this reason, it seems that a range of a few layers, between 3 and 6, suffices 
to reduce most of the undesired NGP scatter.

Extending the pp calculation comes at a price, since the number of operation scales as  $N^{2}$ in the sub-domain. 
This cost is best capture by the increase of real time required by a fixed number of dedicated  {\small CPU}s 
to evolved the particles to the final redshift. Table \ref{table:cpu_pp_ext} presents this usage.

\begin{figure}
  \begin{center}
  \epsfig{file=graphs/pp_ext.eps,width=0.49\textwidth,height=0.49\textwidth}
  \caption{ Dimensionless power spectrum for varying range of the exact pp force, compared to  {\small CAMB} \citep{Lewis:1999bs}.}
    \label{fig:power}
  \end{center}
\end{figure}

\begin{table}
\begin{center}
\caption{Scaling in {\small CPU} resources as a function of the range of the pp interaction.
$N_{layers}$ refers to the number of fine mesh layers around a given cell, inside of which the force calculation
is purely given by the pp contribution. }
\begin{tabular}{|l|c|c|}
\hline 
             Type         & time (h)   \\
                  \hline
PM                         & 1.77 \\
P$^{3}$M             & 2.09 \\
\hline
 $N_{layers}=1$ & 8.74 \\
 $N_{layers}=2$ & 11.47\\
 $N_{layers}=3$ & 14.30 \\
 $N_{layers}=4$ & 18.87\\
 $N_{layers}=5$ & 22.52\\
 $N_{layers}=6$ & 29.62\\
 $N_{layers}=7$ & 34.82\\
 $N_{layers}=8$ & 47.00\\

\hline \hline
\end{tabular}
\label{table:cpu_pp_ext}
\end{center}
\end{table}

Another way to quantify the improvement of the calculation is to look at the halo mass function for these different runs.
{\bf (To do...)}

### 2012-05-01 Introduction.tex
Upgraded extendedpp.tex, plus minor corrections everywhere.

With the advent of large computing facilities, parallel coding has now become common practice, 
and N-body codes have evolved both in performance and complexity. 
Many codes have opted for a tree algorithm \citep{Gadjet, Gadjet2, TPM, GOTPM}, in which 
the local resolution increases with the density of the matter field. 
These have the advantage to balance the load across the computing units, which enable the calculation of high density regions. 
The drawback is a significant loss in speed, which can be only partly recovered by turning off the tree algorithm. 
Unlike Hydra \citep{couchman1991} and the PM code by \cite{FerrelBertschinger1995},
 these are not designed to perform large scale PM calculations.

{\bf (Describe briefly   Hydra, PM code by japanese, and others...)}

{\small PMFAST} (\cite{2005NewA...10..393M}, MPT hereafter) is one of the first code that is designed such as to optimize the PM algorithm,
both in terms of speed and memory usage. It uses a two-level mesh algorithm based on the gravity solver of \cite{TracPen2003},
The long range gravitational force is computed on a  grid four times coarser, such as to minimize the {\small MPI} communication time
and to fit in system's memory. The short range is computed locally on a finer mesh, and only the local sub-volume needs 
to be stored at a given time, allowing for {\small OPENMP}parallel computation.
This this setup enable the code to evolve very large cosmological systems both rapidly and accurately, on relatively modest clusters.
One of the main advantage of {\small PMFAST} over other PM codes is that the number of large arrays is minimized,
and the global {\small MPI} communications are cut down to the minimum: for passing particle at the beginning of each time step,
and  for computing the long range FFTs.
As described in MPT, access to particles is accelerated with the use of linked lists, deletion of `ghost' particles
in buffer zones is done at the same time as particles are passed to adjacent nodes,
and the global FFTs are performed with a slab decomposition of the volumes via a special file transfer interface, 
designed specifically to preserve a high processor load.

>>>>

With the advent of large computing facilities, parallel coding has now become common practice, 
and N-body codes have evolved both in performance and complexity. 
Many codes have opted for a tree algorithm, including Gadjet \citep{Gadjet, Gadjet2}, PMT \citep{1995ApJS...98..355X}, GOTPM \citep{2004NewA....9..111D}, and Hydra \citep{Hydra}, in which the local resolution increases with the density of the matter field. 
These have the advantage to balance the load across the computing units, which enable the calculation of high density regions. 
The drawback is a significant loss in speed, which can be only partly recovered by turning off the tree algorithm. 
The same reasoning applies to the mesh-refine code of  \cite{1991ApJ...368L..23C} and the PM code by \cite{FerrelBertschinger1995}.
 these are not designed to perform large scale PM calculations.

{\bf ( PM code by japanese? Others codes I should mention? That of Hy Trac?)}

{\small PMFAST} (\cite{2005NewA...10..393M}, MPT hereafter) is one of the first code that is designed such as to optimize the PM algorithm,
both in terms of speed and memory usage. It uses a two-level mesh algorithm based on the gravity solver of \cite{2003AAS...203.9703T},
The long range gravitational force is computed on a  grid four times coarser, such as to minimize the {\small MPI} communication time
and to fit in system's memory. The short range is computed locally on a finer mesh, and only the local sub-volume needs 
to be stored at a given time, allowing for {\small OPENMP}parallel computation.
This this setup enable the code to evolve very large cosmological systems both rapidly and accurately, on relatively modest clusters.
One of the main advantage of {\small PMFAST} over other PM codes is that the number of large arrays is minimized,
and the global {\small MPI} communications are cut down to the minimum: for passing particle at the beginning of each time step,
and  for computing the long range FFTs.
As described in MPT, access to particles is accelerated with the use of linked lists, deletion of `ghost' particles
in buffer zones is done at the same time as particles are passed to adjacent nodes,
and the global FFTs are performed with a slab decomposition of the volumes via a special file transfer interface, 
designed specifically to preserve a high processor load.

### 2012-05-01 ScalingPerformance.tex
Upgraded extendedpp.tex, plus minor corrections everywhere.

Real time versus size, bottle neck, extended pp, trials on scinet and ranger, 
p3dfft,

>>>>

Real time versus size, bottle neck, description of the largest runs, etc.
{\bf (Ilian, I will need your input here as well. Hugh, do you have anything you would like to share 
concerning the largest runs you did on Blue Gene? )}

### 2012-05-01 paper.tex
Upgraded extendedpp.tex, plus minor corrections everywhere.

\title[{\small} CUBEP3M: High Performance P3M N-body code]{{\small} CUBEP3M: High Performance P3M N-body code}
\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Hugh Merz, Ilian T. Iliev]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
\thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$ \thanks{E-mail: pen@cita.utoronto.ca}, 
Hugh Merz$^{3}$ \thanks{E-mail: merz@sharcnet.ca} and  Ilian T. Iliev$^{4}$ \thanks{E-mail: i.t.iliev@sussex.ac.uk}\\
%\footnotemark[1]\thanks{This file has been amended to
%highlight the proper use of \LaTeXe\ code with the class file.
%These changes are for illustrative purposes and do not reflect the
%original paper by A. V. Raveendran.}\\
$^{1}$Canadian Institute for Theoretical Astrophysics, University of
Toronto, M5S 3H8, Canada\\
$^{2}$Department of Physics, University of Toronto, M5S 1A7, Ontario,  Canada\\
$^{3}$Department of XXX\\
$^{4}$Department of XXX\\
$^{5}$Department of XXX}

@@

\begin{abstract}
This paper presents {\small CUBEP3M}, an upgraded version of {\small PMFAST}, 
a two-mesh gravity solver that was already among the fastest N-body codes. 
Among the principal changes, the Poisson solver includes particle-particles interactions
at the sub-grid level and across neighbouring cells. The volume decomposition is now cubical,
and most of the calculations on the fine mesh are {\tt openmp} parallel.
Many utilities have been added and are briefly described, including a runtime halofinder
and particle identification tags. 
We discuss the structure of the code, its accuracy, and  its scaling performance.
\end{abstract}

\begin{keywords}
To specify...
\end{keywords}

>>>>

\title[{\small} CUBEP3M: High Performance P$^{3}$M N-body code]{{\small} CUBEP3M: High Performance P$^{3}$M N-body code}
\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Hugh Merz, Ilian T. Iliev]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
\thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$ \thanks{E-mail: pen@cita.utoronto.ca}, 
Hugh Merz$^{3}$ \thanks{E-mail: merz@sharcnet.ca} and  Ilian T. Iliev$^{4}$ \thanks{E-mail: i.t.iliev@sussex.ac.uk}\\
%\footnotemark[1]\thanks{This file has been amended to
%highlight the proper use of \LaTeXe\ code with the class file.
%These changes are for illustrative purposes and do not reflect the
%original paper by A. V. Raveendran.}\\
$^{1}$Canadian Institute for Theoretical Astrophysics, University of
Toronto, M5S 3H8, Canada\\
$^{2}$Department of Physics, University of Toronto, M5S 1A7, Ontario,  Canada\\
$^{3}$Department of XXX\\
$^{4}$Department of XXX\\
$^{5}$Department of XXX}

@@

\begin{abstract}
This paper presents {\small CUBEP3M}, an upgraded version of {\small PMFAST}, 
a two-mesh gravity solver that was already among the fastest N-body codes. 
Among the principal changes, the Poisson solver includes particle-particles interactions
at the sub-grid level and across neighbouring cells, plus each level of the volume decomposition is now cubical.
Force kernel have been improved for enhanced matching of the two mesh contributions near the cutoff length.
We discuss the structure of the code, its accuracy, and its scaling performance.
In addition, many utilities have been added, including a runtime halofinder,
particle identification tags and non-Gaussian initial conditions generator, and we briefly describe their implementation strategy
and accuracy, when applicable .

\end{abstract}

\begin{keywords}
N-body simulations --- Large scale structure of Universe --- Dark matter
\end{keywords}
